---
title: "R Notebook"
output: html_notebook
---

```{r}
dpois(0, 5)
dpois(1, 5)
dpois(2, 5)
dpois(3, 5)
dpois(4, 5)
dpois(5, 5)
dpois(6, 5)
dpois(7, 5)
dpois(8, 5)
dpois(9, 5)
dpois(10, 5)

```
```{r}
pbinom(2, 6, .7)
```
```{r}
pnorm(3.5, 3.8, .2)
pnorm(3.5, 3.8, .2/2)
```
```{r}
pnorm(25, 10)
```
```{r}
#calculating z score
(3.5 - 3.8)/.2
pnorm(-1.5, mean=0, sd=1)
(3.5 - 3.8)/.1
pnorm(-3, mean=0, sd=1)
1- pnorm(60, mean = 50, sd = 13.45)
#Confidence Intervals
750 - qnorm(.025, mean=750, sd = 16)
750 - qnorm(.025, mean=750, sd = 180/sqrt(15))
```
```{r}
#predicting t crit
qt(0.05, df = 11)
qt(.025, df = 29)
qt(.1, df = 17)
#calculating probability of t
pt(-2.45, df = 11)
```
```{r}
#calculating beta 
#find the z crit under the null distribution
Zcrit <- qnorm(0.95, 1496, 129/sqrt(50))
#find probability of overlapping region in real distribution (CDF)
beta <- pnorm(Zcrit, 1532, 129/sqrt(50))
#power is everything in the real distribution that doesn't overlap
pwr <- 1-beta
Zcrit
beta
pwr
```
binomial conditions
• the trials must be independent
• the number of trials, n, must be fixed
• each trial outcome must be classified as a success or failure
• the probability of success, p, must be the same for each trial

conditions for CLT:
Independence: sampled observations must be independent
random sample/assignment
if sampling without replacement, n < 10% of the population
Sample size/skew: either the population is normal, or if the distribution is skewed, the sample size is large (n > 30)

regardless of the parent population…

#1: the mean of the population of means is always equal to the mean of the parent population
#2: the standard deviation, s, of the population of means is always equal to the standard deviation of the parent population, σ,  divided by the square root of the sample size
#3: the distribution of means will approximate a normal distribution as the sample size, n, of samples increases




```